thesis2024

### 2025-05-07 01-intro.qmd
Abstract and other elements of thesis. Pre-render corrections

Some studies suggest a chronological correlation between climatic shifts and the emergence of major technocomplexes, implying that cultural reorganization may have been triggered by environmental changes that altered landscapes and essential resources [e.g., @bradtmoller_repeated_2012-1; @cascalheiraHunterGathererEcodynamics2013; @bicho_rapid_2017-1]. These studies also indicate shifts in territory expansion/contraction, settlement types, mobility, and intergroup contacts [@schmidtRapidClimateChange2012]. While cultural changes are not necessarily deterministic responses to climate fluctuations, it is evident that hunter-gatherers adapted to their landscapes. Cultural, technological, and social transformations provide key insights into how and why human groups adapted at different times.

@@

A clear example is the identification of long-distance contacts between regions in Iberia, suggesting complex mobility and procurement patterns throughout the UP [e.g., @hermidaSilexNWPeninsula2016; @corchonrodriguezMobiliteTerritoiresRelations2016; @ortegaMobilitatDesplacamentsDels2003; @sanchezdelatorreNewDataChert2023]. The Côa Valley in northeastern Portugal is a particularly emblematic case, as non-local chert is present throughout the UP sequence, with potential sources located 150–250 km away in both southwestern (central Portugal) and southeastern (central Spain) directions. These patterns have been interpreted as evidence of exchange networks facilitated by seasonal movements and social interactions between groups across Iberia [@aubryFarFlintInferring2022; @aubryRawMaterialProcurement2004].

@@

Vale Boi, located in southwestern Iberia, is a multi-component site spanning several UP technocomplexes, from the Gravettian (c. 32 ka cal BP) to the Magdalenian [c. 15 ka cal BP; @bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site, considered an eco-cultural niche, provided stable climatic conditions even during colder periods, supporting continuous residential occupations that have been consistently dated [@bicho_ecodynamics_2013; @cascalheira_2017; @belmiro_gravettian-solutrean_2021]. Extensive lithic assemblages from Vale Boi have undergone technological analysis, revealing patterns of technological change and resilience [@belmiro_gravettian-solutrean_2021; @cascalheiraValeBoiAlgarve2013; @gibajabaoProvenienceTechnologyMorphology2013; @horta_role_2019; @marreirosLithicTechnologyGravettian2015]. The site has also been identified as a centre for knowledge transfer, particularly during the Last Glacial Maximum [@cascalheiraINFLUENCIAMEDITERRANICANAS; @cascalheiraTerritorialityOrganizationTechnology2019].

@@

-   [@sec-chapter5], titled "From Stone to Tool: How Raw Materials Influenced Upper Palaeolithic Technology at Vale Boi, Southwestern Iberia," combines the previously obtained data on chert types (particularly regarding the distinction between local and non-local cherts) with technological data. The paper explores various patterns of chert use and management.

>>>>

Some studies suggest a chronological correlation between climatic shifts and the emergence of major technocomplexes, implying that cultural reorganization may have been triggered by environmental changes that altered landscapes and essential resources [e.g., @bicho_rapid_2017-1; @bradtmoller_repeated_2012-1; @cascalheiraHunterGathererEcodynamics2013]. These studies also indicate shifts in territory expansion/contraction, settlement types, mobility, and intergroup contacts [@schmidtRapidClimateChange2012]. While cultural changes are not necessarily deterministic responses to climate fluctuations, it is evident that hunter-gatherers adapted to their landscapes. Cultural, technological, and social transformations provide key insights into how and why human groups adapted at different times.

@@

A clear example is the identification of long-distance contacts between regions in Iberia, suggesting complex mobility and procurement patterns throughout the UP [e.g., @corchonrodriguezMobiliteTerritoiresRelations2016; @hermidaSilexNWPeninsula2016; @ortegaMobilitatDesplacamentsDels2003; @sanchezdelatorreNewDataChert2023]. The Côa Valley in northeastern Portugal is a particularly emblematic case, as non-local chert is present throughout the UP sequence, with potential sources located 150–250 km away in both southwestern (central Portugal) and southeastern (central Spain) directions. These patterns have been interpreted as evidence of exchange networks facilitated by seasonal movements and social interactions between groups across Iberia [@aubryFarFlintInferring2022; @aubryRawMaterialProcurement2004].

@@

Vale Boi, located in southwestern Iberia, is a multi-component site spanning several UP technocomplexes, from the Gravettian (c. 32 ka cal BP) to the Magdalenian (c. 15 ka cal BP) [@bicho_ecodynamics_2013; @bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site, considered an eco-cultural niche, provided stable climatic conditions even during colder periods, supporting continuous residential occupations that have been consistently dated [@bicho_ecodynamics_2013; @cascalheira_2017; @belmiro_gravettian-solutrean_2021]. Extensive lithic assemblages from Vale Boi have undergone technological analysis, revealing patterns of technological change and resilience [@belmiro_gravettian-solutrean_2021; @cascalheiraValeBoiAlgarve2013; @gibajabaoProvenienceTechnologyMorphology2013; @horta_role_2019; @marreirosLithicTechnologyGravettian2015]. The site has also been identified as a centre for knowledge transfer, particularly during the Last Glacial Maximum [@cascalheiraINFLUENCIAMEDITERRANICANAS; @cascalheiraTerritorialityOrganizationTechnology2019].

@@

-   [@sec-chapter5], titled "From Stone to Tool: How Raw Materials Influenced Upper Palaeolithic Technology at Vale Boi, Southwestern Iberia," (submitted) combines the previously obtained data on chert types (particularly regarding the distinction between local and non-local cherts) with technological data. The paper explores various patterns of chert use and management.
